New and simplified manual controls for projection and slice tours, with application to exploring classification boundaries in high dimensions

### 2023-01-19 review/response.Rmd

---
title: "Response to review: New and simplified manual controls for projection and slice tours, with application to exploring classification boundaries in high dimensions"
author: "Ursula Laa, Alex Aumann, Dianne Cook, German Valencia"
date: "2023-01-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

 XXX If use new slice center control for fig 7, need to make a note of it here, updated the mathematica notebook and check text and captions correct.

 XXX References need to be consistent, all initials or all full names, and all sentence case or title case on titles.

@@

Section 1 line 29 "The software section describes mathematica package"  mathematica should be capitalized as you did  through the paper in most cases.  Also since is the first time you have mentioned Mathematica you should include the reference here (Wolfram reference appears in page 4 line 26 for first time)

\textcolor{blue}{Done}

Section 2: "manual tour allows the user to alter the coefficients of one (or more) variables" but in the explanation it is not clear what is  $V_m$  and how to alter the coefficients for more than one variable.

\textcolor{blue}{Definition of $V_m$ added and Figure 1 has been annotated to make the algorithm easier to understand. Controlling more than one variable is possible with the R implementation and this has been noted in the text.}

@@

In page 4 line 2.2 "The components corresponding to $V_m$ are directly controlled by cursor movement which updates row $m$ of $A$". Maybe a video will be illustrative in this sub-section.

\textcolor{blue}{Interpreting this to be Section 2.2, not line 2.2. Annotation has been added to existing figure.}

@@

Section 3: will be useful to include a schematic diagram illustrating the interactivity

\textcolor{blue}{We're not sure what this refers to. The additional details and explanations of the algorithm are provided in the appendix. Additional refinement, and figure annotations, were made in the previous sections.}

@@

In the manual tour implementation.
This part is difficult to follow, it is more clear after reading the application Section because you have included some videos but I think you need to include something here to help the reader.

\textcolor{blue}{Additional explanation, clearer notation, and annotation of the plot has been done to help here.}

@@

In the introduction "The software section describes a mathematica package that is used for the application, and describes the interactive environment that would be desirable within R as new technology becomes available"
I think you need to be more specific about the R limitations.
Iin Section 3. Page 7, line 20 "One might implement such an interactive interface via R Shiny, but in particular the tracking of the mouse position in small increments might pose a challenge" This  is the only possible R limitation you have mentioned.

\textcolor{blue}{Additional text has been added to explain why this is difficult in R/Shiny/rgl.}

### 2023-01-19 paper/paper.Rmd
handling various reviewer requests

<!-- Background -->
From a statistical perspective 3D is a rare data dimension, so unlike in most 3D rotation computer graphics applications, \textcolor{blue}{methods which work for arbitrary dimensions have more general utility}. A good approach is to show projections from an arbitrary dimensional space to create dynamic data visualizations called *tours*. Tours involve views of high-dimensional ($p$) data with low-dimensional ($d$) projections. In his original paper on the grand tour, @As85 provided several algorithms for tour paths that could theoretically show the viewer the data *from all sides*. Prior to Asimov's work, there were numerous preparatory developments including @tukey's PRIM-9. PRIM-9 had user-controlled rotations on coordinate axes, allowing one to manually tour through low-dimensional projections. (A video illustrating the capabilities is available through \textcolor{blue}{the} video library of @ASA22.) Steering through all possible projections is impossible, unlike Asimov's tours which allows one to quickly see many, many different projections. After Asimov there have been many tour developments, which are summarized in @lee2021.

@@

A manual tour allows the user to alter the coefficients of one (or more) variables contributing to a $d$ dimensional projection. The initial ingredients are an orthonormal basis ($A_{p\times d}$) defining the projection of the data, and a variable id ($m \in \{1, ..., p\}$) specifying which coefficient will be \textcolor{blue}{changed: the $m^{th}$ row of $A_{p\times d}$ which we denote by $V_m$.} A method to update the values of the component of the controlled variable $V_m$ is then needed.

@@

This algorithm will produce the changes to a projection as illustrated in Figure \ref{fig:manualsequence}. The controlled variable, $V_m$, corresponds to the black line, and sequential changes to row $m$ of $A$ can be seen to roughly follow a specified position (orange dot). Changes in the other components happen as a result of the orthonormalisation, but are uncontrolled.

```{r manualsequence, out.width="100%", width=12, height=4, fig.cap="Sequence of projections where contribution of one variable is controlled (black) is changed using unconstrained orthonormalisation. The dot (orange) indicates the chosen values for the controlled variable. It can be seen that the actual axis does not precisely match the chosen position, but it is close."}
source("../code/plot_basis.R")
source("../code/linear_alg.R")

@@

# Now set a new position
vchange <- 3
eps <- 0.1
plot_basis(A, vchange, vcol="black")

# Iterate and save images
# Unconstrained orthonormalisation
Anew <- A
for (i in 1:3) {
  Anew[vchange,] <- A[vchange,]*(1-i*eps)
  exact <- Anew[vchange,]
  Anew <- tourr::orthonormalise(Anew)
  plot_basis(Anew, vchange, vcol="black")
  points(exact[1], exact[2], pch=16, col="orange")
}

@@

\textcolor{blue}{The Mathematica implementation relies on the inbuilt Mathematica function  `LocatorPane`. This function creates a region on the screen where the position of the mouse is captured and then converted to input that updates the graphics functions, enabling the manual navigation of the tour. We then use the graphics function `ListPlot` embedded in dynamic objects via  `DynamicModule` to produce the display. Control objects, such as sliders, locator panes, and input fields, are used on dynamic variables, and when there is a change in the dynamic variable the dynamic objects which contain that variable will be updated.}

@@

The R package `tourr` provides numerous types of tours. An additional function, `radial_tour()` has been constructed that will generate a sequence of projections that will decrease the coordinates for $V_m$ to $\boldsymbol{0}$ and back to the original values. It is applicable for any projection dimension, $d$. Changing slice center manually can be accomplished with the new function, `manual_slice()`. This changes the value of the center point ($\mathbf{c}$) of the slice, along a selected variable axis. It slices from the center of the data out to one side, back to the center, out to the opposite side, and then back to the center. The appendix contains a discussion of a possible visual guide for the slice center location.

@@

- The Mathematica source code defining the new functions in mmtour.wl.
- Three Mathematica notebooks with the code used in the application (corresponding to the three subsections).
- Appendix with additional details about manual controls, and Mathematica functions.
- Animations illustrating the manual tour and slicing, matching the static figures in the paper. These are also available at https://vimeo.com/747585410 and https://vimeo.com/747590472.

>>>>

<!-- Background -->
From a statistical perspective 3D is a rare data dimension, so unlike in most 3D rotation computer graphics applications, \textcolor{blue}{methods which work for arbitrary dimensions have more general utility}. A good approach is to show projections from an arbitrary dimensional space to create dynamic data visualizations called *tours*. Tours involve views of high-dimensional  data with low-dimensional projections. In his original paper on the grand tour, @As85 provided several algorithms for tour paths that could theoretically show the viewer the data *from all sides*. (Various videos of tours, e.g. @dataviewer, @gt-pp-video, on the @ASA22 illustrate this concept.) Prior to Asimov's work, there were numerous preparatory developments including @tukey's PRIM-9. PRIM-9 had user-controlled rotations on coordinate axes, allowing one to manually tour through low-dimensional projections. (A video [@PRIM9-video] illustrating the capabilities is available through \textcolor{blue}{the} video library of @ASA22.) Steering through all possible projections is impossible, unlike Asimov's tours which allows one to quickly see many, many different projections. After Asimov there have been many tour developments, which are summarized in @lee2021.

@@

A manual tour allows the user to alter the coefficients of one (or more) variables, of $p$, contributing to a $d$-dimensional projection. The initial ingredients are an orthonormal basis ($A_{p\times d}$) defining the projection of the data, and a variable id ($m \in \{1, ..., p\}$) specifying which coefficient will be \textcolor{blue}{changed: the $m^{th}$ row of $A_{p\times d}$ which we denote by $V_m$.} A method to update the values of the component of the controlled variable $V_m$ is then needed.

@@

This algorithm will produce the changes to a projection as illustrated in Figure \ref{fig:manualsequence}. The controlled variable, $V_m$, corresponds to the dark line, and sequential changes to row $m$ of $A$ can be seen to roughly follow a specified position (dot). Changes in the other components happen as a result of the orthonormalisation, but are uncontrolled.

The algorithm will also allow for more than one variable to be controlled. This is more difficult to implement interactively, but is easy with scripting, as is available in the R implementation.

```{r manualsequence, out.width="100%", width=12, height=4, fig.cap="Sequence of projections where contribution of one variable is controlled (dark) is changed using unconstrained orthonormalisation. The dot indicates the chosen values for the controlled variable, $V_m$. It can be seen that the actual axis does not precisely match the chosen position, but it is close."}
source("../code/plot_basis.R")
source("../code/linear_alg.R")

@@

# Now set a new position
vchange <- 3
eps <- 0.1
plot_basis(A, vchange, vcol="black")
text(A[vchange, 1], A[vchange, 2], expression(V[m]), pos=3, col="orange")
points(A[vchange, 1], A[vchange, 2], pch=16, col="orange")

# Iterate and save images
# Unconstrained orthonormalisation
Anew <- A
for (i in 1:3) {
  Anew[vchange,] <- A[vchange,]*(1-i*eps)
  exact <- Anew[vchange,]
  Anew <- tourr::orthonormalise(Anew)
  plot_basis(Anew, vchange, vcol="black")
  text(Anew[vchange, 1], Anew[vchange, 2], expression(V[m]), pos=3, col="orange")
  points(exact[1], exact[2], pch=16, col="orange")
}

@@

\textcolor{blue}{The Mathematica implementation relies on the inbuilt function  `LocatorPane`. This function creates a region on the screen where the position of the cursor is captured and then converted to input that updates the graphics functions, enabling the manual navigation of the tour. We then use the graphics function `ListPlot` embedded in dynamic objects via  `DynamicModule` to produce the display. User interface objects, such as sliders, locator panes, and input fields, are used to provide user control of slice thickness, center, scale and point size.}

@@

The R package `tourr` provides numerous types of tours, displayed as animations.

The new function, `radial_tour()` has been constructed that will generate a sequence of projections that will decrease the coordinates for $V_m$ to $\boldsymbol{0}$ and back to the original values. It is applicable for any projection dimension, $d$. The script interface has one advantage over an interactive interface, in that two or more variables can be controlled simultaneously.

Changing slice center manually can be accomplished with the new function, `manual_slice()`. This changes the value of the center point ($\mathbf{c}$) of the slice, along a selected variable axis. It slices from the center of the data out to one side, back to the center, out to the opposite side, and then back to the center. <!--The appendix contains a discussion of a possible visual guide for the slice center location.-->

@@

- The Mathematica source code defining the new functions in mmtour.wl.
- Three Mathematica notebooks with the code used in the application (corresponding to the three subsections).
- Appendix with additional details about manual controls, and Mathematica functions.
- The `Rmd` file that generated the paper, containing the R code to produce Figure 1.
- Animations illustrating the manual tour and slicing, matching the static figures in the paper. These are also available at https://vimeo.com/747585410 and https://vimeo.com/747590472.

### 2023-01-19 paper/paper.tex
handling various reviewer requests

From a statistical perspective 3D is a rare data dimension, so unlike in
most 3D rotation computer graphics applications,
\textcolor{blue}{methods which work for arbitrary dimensions have more general utility}.
A good approach is to show projections from an arbitrary dimensional
space to create dynamic data visualizations called \emph{tours}. Tours
involve views of high-dimensional (\(p\)) data with low-dimensional
(\(d\)) projections. In his original paper on the grand tour,
\citet{As85} provided several algorithms for tour paths that could
theoretically show the viewer the data \emph{from all sides}. Prior to
Asimov's work, there were numerous preparatory developments including
\citet{tukey}'s PRIM-9. PRIM-9 had user-controlled rotations on
coordinate axes, allowing one to manually tour through low-dimensional
projections. (A video illustrating the capabilities is available through
\textcolor{blue}{the} video library of \citet{ASA22}.) Steering through
all possible projections is impossible, unlike Asimov's tours which
allows one to quickly see many, many different projections. After Asimov
there have been many tour developments, which are summarized in
\citet{lee2021}.

@@

A manual tour allows the user to alter the coefficients of one (or more)
variables contributing to a \(d\) dimensional projection. The initial
ingredients are an orthonormal basis (\(A_{p\times d}\)) defining the
projection of the data, and a variable id (\(m \in \{1, ..., p\}\))
specifying which coefficient will be
\textcolor{blue}{changed: the $m^{th}$ row of $A_{p\times d}$ which we denote by $V_m$.}
A method to update the values of the component of the controlled
variable \(V_m\) is then needed.

@@

This algorithm will produce the changes to a projection as illustrated
in Figure \ref{fig:manualsequence}. The controlled variable, \(V_m\),
corresponds to the black line, and sequential changes to row \(m\) of
\(A\) can be seen to roughly follow a specified position (orange dot).
Changes in the other components happen as a result of the
orthonormalisation, but are uncontrolled.

\begin{figure}
\includegraphics[width=1\linewidth]{paper_files/figure-latex/manualsequence-1} \caption{Sequence of projections where contribution of one variable is controlled (black) is changed using unconstrained orthonormalisation. The dot (orange) indicates the chosen values for the controlled variable. It can be seen that the actual axis does not precisely match the chosen position, but it is close.}\label{fig:manualsequence}
\end{figure}

@@

The R package \texttt{tourr} provides numerous types of tours. An
additional function, \texttt{radial\_tour()} has been constructed that
will generate a sequence of projections that will decrease the
coordinates for \(V_m\) to \(\boldsymbol{0}\) and back to the original
values. It is applicable for any projection dimension, \(d\). Changing
slice center manually can be accomplished with the new function,
\texttt{manual\_slice()}. This changes the value of the center point
(\(\mathbf{c}\)) of the slice, along a selected variable axis. It slices
from the center of the data out to one side, back to the center, out to
the opposite side, and then back to the center. The appendix contains a
discussion of a possible visual guide for the slice center location.

@@

\begin{itemize}
\tightlist
\item
  The Mathematica source code defining the new functions in mmtour.wl.
\item
  Three Mathematica notebooks with the code used in the application
  (corresponding to the three subsections).
\item
  Appendix with additional details about manual controls, and
  Mathematica functions.
\item
  Animations illustrating the manual tour and slicing, matching the
  static figures in the paper. These are also available at
  \url{https://vimeo.com/747585410} and
  \url{https://vimeo.com/747590472}.
\end{itemize}

>>>>

From a statistical perspective 3D is a rare data dimension, so unlike in
most 3D rotation computer graphics applications,
\textcolor{blue}{methods which work for arbitrary dimensions have more general utility}.
A good approach is to show projections from an arbitrary dimensional
space to create dynamic data visualizations called \emph{tours}. Tours
involve views of high-dimensional data with low-dimensional projections.
In his original paper on the grand tour, \citet{As85} provided several
algorithms for tour paths that could theoretically show the viewer the
data \emph{from all sides}. (Various videos of tours, e.g.
\citet{dataviewer}, \citet{gt-pp-video}, on the \citet{ASA22} illustrate
this concept.) Prior to Asimov's work, there were numerous preparatory
developments including \citet{tukey}'s PRIM-9. PRIM-9 had
user-controlled rotations on coordinate axes, allowing one to manually
tour through low-dimensional projections. (A video \citep{PRIM9-video}
illustrating the capabilities is available through \textcolor{blue}{the}
video library of \citet{ASA22}.) Steering through all possible
projections is impossible, unlike Asimov's tours which allows one to
quickly see many, many different projections. After Asimov there have
been many tour developments, which are summarized in \citet{lee2021}.

@@

A manual tour allows the user to alter the coefficients of one (or more)
variables, of \(p\), contributing to a \(d\)-dimensional projection. The
initial ingredients are an orthonormal basis (\(A_{p\times d}\))
defining the projection of the data, and a variable id
(\(m \in \{1, ..., p\}\)) specifying which coefficient will be
\textcolor{blue}{changed: the $m^{th}$ row of $A_{p\times d}$ which we denote by $V_m$.}
A method to update the values of the component of the controlled
variable \(V_m\) is then needed.

@@

This algorithm will produce the changes to a projection as illustrated
in Figure \ref{fig:manualsequence}. The controlled variable, \(V_m\),
corresponds to the dark line, and sequential changes to row \(m\) of
\(A\) can be seen to roughly follow a specified position (dot). Changes
in the other components happen as a result of the orthonormalisation,
but are uncontrolled.

The algorithm will also allow for more than one variable to be
controlled. This is more difficult to implement interactively, but is
easy with scripting, as is available in the R implementation.

\begin{figure}
\includegraphics[width=1\linewidth]{paper_files/figure-latex/manualsequence-1} \caption{Sequence of projections where contribution of one variable is controlled (dark) is changed using unconstrained orthonormalisation. The dot indicates the chosen values for the controlled variable, $V_m$. It can be seen that the actual axis does not precisely match the chosen position, but it is close.}\label{fig:manualsequence}
\end{figure}

@@

The R package \texttt{tourr} provides numerous types of tours, displayed
as animations.

The new function, \texttt{radial\_tour()} has been constructed that will
generate a sequence of projections that will decrease the coordinates
for \(V_m\) to \(\boldsymbol{0}\) and back to the original values. It is
applicable for any projection dimension, \(d\). The script interface has
one advantage over an interactive interface, in that two or more
variables can be controlled simultaneously.

Changing slice center manually can be accomplished with the new
function, \texttt{manual\_slice()}. This changes the value of the center
point (\(\mathbf{c}\)) of the slice, along a selected variable axis. It
slices from the center of the data out to one side, back to the center,
out to the opposite side, and then back to the center.

@@

\begin{itemize}
\tightlist
\item
  The Mathematica source code defining the new functions in mmtour.wl.
\item
  Three Mathematica notebooks with the code used in the application
  (corresponding to the three subsections).
\item
  Appendix with additional details about manual controls, and
  Mathematica functions.
\item
  The \texttt{Rmd} file that generated the paper, containing the R code
  to produce Figure 1.
\item
  Animations illustrating the manual tour and slicing, matching the
  static figures in the paper. These are also available at
  \url{https://vimeo.com/747585410} and
  \url{https://vimeo.com/747590472}.
\end{itemize}
